Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2024-01-16 chapters/Method.tex

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes.}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we cannot find an analytical solution to every given flow problem, which is due to the shape and the structure of the Navier-Stokes equations.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille flow'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{suteraHistoryPoiseuilleLaw}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number\footnote{The Stokes number is a dimensionless number that defines the behaviour of particles suspended in a flow and can be calculated with $St = \frac{t_0u_0}{l_0}$, where $t_0$ is a relaxation time of the particles, $u_0$ is a characteristic velocity and $l_0$ is a characteristic length scale.}.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
But going from laminar to turbulent flows with high Reynolds numbers poses still a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Several physics problems are being studied to large extent with numerical tools, quantum field theory~\cite{montvayQuantumFieldsLattice1994}, solid mechanics~\cite{cardiffThirtyYearsFinite2021}, and as outlined in the introduction Chap.~\ref{chapter:intro} computational fluid dynamics (CFD) is one of those fields.
Although chip manufacturers have a hard time to keep up with Moore's law, computing power is still growing rapidly~\cite{schallerMooreLawPresent1997}.
Of course instead of ever faster processors (measured in clock rate), the trend goes in the direction of parallel and accelerated computing. 
Several accelerator designs hit the market around 2020, one design that works particularly well as accelerators are graphics processing units (GPUs).
While a GPU lacks the complexity of a CPU it excels at doing simple task over and over again.
For example the generation of colorful pixels for a monitor.
It turns out that several algorithms developed for solving differential equations are well suited to be calculated on the GPU. 
Among them is the lattice Boltzmann method, at least a very basic implementation of the method.
GPUs however are not the only devices to be used as accelerators.
Intel for example used a different approach to accelerator computing and already some time ago developed the Xenon Phi.
However with lack of performance as compared to Nvidia's graphics cards and company intern decisions Intel ended the Xenon Phi project in 2020.
The successor of the Xenon Phi project is the very recent \textit{oneAPI} approach and its hardware~\cite{christgauPortingLegacyCUDA2020}.
Shifting the focus towards heterogeneous compute architectures, or as Rick Stevens\footnote{Rick Stevens is Professor at the Department of Computer Science and Associate Laboratory Director, Computing, Environment and Life Sciences at Argonne National Laboratory.} said: ''The future of advanced computing requires heterogeneous hardware to maximize the computing power needed for exascale-class workloads. The oneAPI industry initiative Intel is spearheading will ensure that programming across diverse compute architectures is greatly simplified.''
Not only is this Intel's idea but a general trend in high performance computing (HPC) moving from CPU only clusters to heterogeneous hardware.
Apart from Intel more and more companies joined the development of workload optimized hardware.
Which is possible not only but also due to the technological inventions from arm (Advanced RISC Machines) and other system on a chip (SoC) design companies.
SoC designs optimized for specific workloads is the key business of arm, which is a paradigm change of the industry.
For the longest time established chip manufactures also defined the designs of a chips.
Today SoC designs can be manufactured at project manufacturers with TSMC (Taiwan Semiconductor Manufacturing Company Limited) being one of them.
TSMC is also the world leader in the production of semiconductors for computer chips. 
They moved the resolution limit for single transistors from 12nm to about 5nm in roughly 5 years.
These technological steps allowed on the other hand companies such as Apple to design SoCs for their specific needs. 
The M1 Max is from this perspective a very interesting device, although it is not an x86 design.

@@

\section{Numerical Methods}
\label{sec:numerical_methods}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{graphics/800px-Finite_Differences.svg.png}
    \caption{Discretization of a continuous function y (in red) along a discrete space x ($x_0, x_1, x_2$,...).
    Consecutive points on the x-axis are separated by a uniform distance $h$.
    }
    \label{fig:finite_difference}
\end{figure}
The lattice Boltzmann method is one among many to simulate the behaviour of a fluid.
Especially when it comes to numerical simulations of the thin film equation, see Eq.~(\ref{eq:simple_thin_film}), other numerical approaches are more established and have been successfully applied for more than thirty years~\cite{beckerComplexDewettingScenarios2003, peschkaSignaturesSlipDewetting2019, davidovitchSpreadingViscousFluid2005, meckeThermalFluctuationsThin2005, diezGlobalModelsMoving2000, schwartzSimulationDropletMotion1998}.
The thin film equation is a non-linear (stiff) fourth order PDE which requires choosing an appropriate scheme, because not every numerical scheme will be able to provide first stability and second a correct approximation of the solution.
A numerical scheme is a blue print of how to discretize derivatives and arrange terms such that an iterative process converges towards a solution.
That said, developing new numerical schemes for PDEs is still a very active research area.
% In a time dependent problem, which the thin film equation is, we have to be aware that a discretization in space may be very different from a discretization in time.
% Often a good approach is to first solve the homogenous problem and only than advance in time.
One way to compute this differential operators on a computer is via finite differences.

\subsection{Finite differences}
Using finite differences means nothing more than discretizing the continuous problem into small (infinitesimal) discrete segments.
Figure~\ref{fig:finite_difference} illustrates a discretization approach where the function $f(x)$ is approximated by a finite set of points which are shown as black dots on the red line.
At every point $x_i$ along the x-axis we know the value of the function $f(x_i)$.
This however does not mean that we know the value of $f(x)$ at arbitrary $x$ values and even worse we do not know $\partial_x f(x_i)$ at $x_i$. 
For a continuous function, e.g. the film thickness $h(\mathbf{x},t)$, we can analytically compute the derivatives. 
Here we are left with a discrete set of point and have yet to come up with an idea to calculate $\nabla h(\mathbf{x}_i, t)$, which is crucial to compute the film pressure, Eq.~(\ref{eq:pfilm}).

Discretization of a derivative is a long studied problem~\cite{booleTreatiseCalculusFinite1872, jordanCalculusFiniteDifferences1965}.
The Taylor expansion gives us a tool to approximate the solution to a differential.
In one spatial dimension there are three different methods to compute the derivative, neglecting boundary conditions there is
\begin{equation}\label{eq:forward_dif}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_i)}{h},
\end{equation}
where $h$ is the spacing between two consecutive points as introduced in Fig.~\ref{fig:finite_difference}. 
Subtracting the value of the function $f(x_{i+1})$ from $f(x_i)$ is called forward difference.
Changing the interval the backward difference is given as
\begin{equation}\label{eq:backward_dif}
    \partial_x f(x_i) = \frac{f(x_i) - f(x_{i-1})}{h}.
\end{equation}
Both the forward as well as the backward differential are first order in their error, beyond first order is the central difference approach
\begin{equation}\label{eq:center_diff}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_{i-1})}{2h},
\end{equation}
where the value of the derivative is averaged over a larger interval. 
The second derivative is than simply applying the differential operator (forward, backward, central) to  the first derivative, in case of central difference this yields
\begin{equation}\label{eq:second_central_diff}
    \partial_x^2 f(x_i) = \frac{f(x_{i+1})-2f(x_i)+f(x_{i-1})}{^2}.
\end{equation}

However all finite difference schemes suffer from the concrete choice of the discretization, which is often referred as discretization artifacts.
The starting equation is usually a continuous one in both space and time, but the numerical solver requires a discretized system. 
This can be done as shown in Fig.~\ref{fig:finite_difference} by introducing a grid for the spatial dimensions. 
For now we assume that we can the same to the time and put the temporal dimension into a discrete grid. 
To understand the error one makes using finite difference we use the example of a sine wave. 
We mark two points along the curve and connect them, which yields a straight line. 
Therefore all information in between the points is lost, instead of $y(x) = \sin(x)$ one would guess $y(x) = kx + d$ for the points.
Besides the fact that information about the function is lost it still could be a good enough approximation for the derivative, especially if the points are not too far apart.
Increasing the number of points along the curve, thus making $h$ smaller, will increase the accuracy of the resulting approximation as displayed in Fig.~\ref{fig:finite_difference}.
A good approximation for $y(x) = \sin(x)$ requires therefore more than just two points.
Which explains the so called fidelity of the finite difference method~\cite{guruswamyReviewNumericalFluids2002}.
One needs a fine enough grid to resolve the to be studied dynamics, but the smaller the spacing $h$ gets the more numerically demanding is the calculation.

In fact there are further limitations to the problem of numerically approximating a PDE.
Numerical methods for PDEs, e.g. the thin film or the wave equation, need to satisfy the CFL condition~\cite{courantUeberPartiellenDifferenzengleichungen1928}.  
Beyond other things it states that choosing an increment for time cannot be independent of the spatial resolution.
This condition can be formalized as
\begin{equation}\label{eq:CFL}
    C = \Delta t \left(\sum_{i=1}^n \frac{u_{x_i}}{\Delta x_i}\right) \leq C_{max},
\end{equation}
with $\Delta x_i$ and $\Delta t$ being the spatial and temporal increment, respectively and $u_{x_i}$ being the magnitude of the velocity in the respective direction.
In case of Fig.~\ref{fig:finite_difference} we have $\Delta x = h$. 
Behind this concept is the idea that propagation of information has an upper limit.
For fluid dynamic problems this upper limit is the speed of sound, in electrodynamics it would be the speed of light.
Any propagation larger than this upper limit introduces stability issues.
These issues are purely numerical as we know that air can flow with velocities larger than the speed of sound, however by doing so the flow creates a so called shock.

Thinking about a wave package, numerically it should be ensured that several time iterations are computed before the package has reached the next spatial point of the grid.
The maximal value that the Courant number can reach for an explicit solver, to which for example the lattice Boltzmann method belongs (assuming a BGK collision operator), is $C_{max} = 1$.
For implicit methods the maximal Courant number can be larger than one.
% In Fig.~\ref{fig:finite_difference} there is a single lattice constant $h$ that sets the spatial resolution of the system as such $\Delta x = h$.
The iteration from $u(x_i,t_0)$ to $u(x_i,t_1)$ is called a time step. 
The ``distance'' between consecutive time values, here $\Delta t = t_1 - t_0$ must ensure that forces and therefore acceleration are well within the stability condition Eq.~(\ref{eq:CFL}).

@@

\subsection{Spectral methods}
Another strategy to solve a differential equation is to perform a transformation.
One rather famous transformation is the Fourier Transformation (FT), which we have used to find the dispersion relation in Sec.~\ref{sec:thin_films} and will encounter again in Chap.~\ref{chapter:second_paper}.
The set of solvers using a FT to solve a given differential equation is called spectral methods.
Doing so has the advantage of interchanging the derivatives with simple multiplications.
For example the term $\partial_x f(x)$ transforms to  
\begin{equation}\label{eq:fourier_transform}
    \partial_x f(x) = q\cdot \tilde{f}(q),
\end{equation}
where $\tilde{f}(q)$ is the Fourier transformed of $f(x)$, e.g. $\tilde{f}(q) = \delta(q - a/2\pi)$ for $f(x) = e^{iax}$. 
Instead of solving a differential equation one solves an algebraic equation, or a set of algebraic equations.
While it is often straightforward to perform the transformation it can become problematic to perform the backward transformation. 
Usually the expressions are fairly complicated and often the backward transformation into real space is not easy to perform.
In Chap.~\ref{chapter:second_paper} we arrive at such an expression for the structure factor $\mathcal{S}(q, t)$.
A structure factor is a measure of the distribution of length scales, roughly speaking.
In this specific case there is no need for a backward transformation because (scattering) experiments usually supply data in Fourier space.

However, when simulating the thin film equation we may be interested in forces, pressures or simply the thickness $h(\mathbf{x},t)$, which is why we need a backward transformation.
Numerically speaking the operators for transformations are matrices, which depend on the explicit problem.
Solutions in real space are generated by inversion of these operators and therefore by inversion of an (arbitrary) matrix.
Computing the inverse of a matrix is a demanding problem. 
Nevertheless, these transformation approaches are often quite helpful finding an exact solution.
One rather prominent example is the computation of Green's functions, e.g. in electrostatics~\cite{greenEssayApplicationMathematical1889}.
For the shallow water theory, see Sec.~\ref{sec:theory_shallow_water}, one class of effective methods are the (discontinuous) Galerkin methods~\cite{ernTheoryPracticeFinite2004} and to some extent the related finite element method (FEM).

Literature concerning simulations of the thin film equation using spectral solvers is rather sparse. 
That does however not mean that spectral methods are ill suited for this specific problem.
Dur{\'a}n-Olivencia, M. et al. have shown that a spectral method is not only able to simulate the thin film equation but also they also included thermal fluctuations and were able to simulate the stochastic thin film equation~\cite{duran-olivenciaInstabilityRuptureFluctuations2019}.

\subsection{Molecular dynamics}
The third approach we discuss in this section is the molecular dynamics method (MD)~\cite{haileMolecularDynamicsSimulation1997, zhangMolecularSimulationThin2019, wengMolecularDynamicsInvestigation2000}.
When we solve the equations of motion of molecules we do not per se think of a \textbf{thin} film flow, however molecular dynamics can be very insightful when dealing with thin film problems. 
In Sec.~\ref{sec:thin_films} we introduced the term wetting and the three phase contact line.
Advancement of the three phase contact line can be boiled down to the interaction between fluid molecules and substrate molecules, therefore MD is perfectly suited to study e.g. slip~\cite{jabbarzadehWallSlipMolecular1999, segaRegularizationSlipLength2013}. 
Instead of working with macroscopic fields, such as density $\rho$ or velocity $\mathbf{u}$, the method uses a bottom up approach.
The main idea is to compute the trajectory of every particle using Newton's equation of motion, therefore to assume that particles are classical objects.
Interactions as well as boundary conditions can be added via pair potentials.
A common interaction potential is the Lennard-Jones potential
\begin{equation}\label{eq:Len-Jon}
    U(r_{ij}) = 4\varepsilon\left[\left(\frac{\sigma}{r_{ij}}\right)^{12}-\left(\frac{\sigma}{r_{ij}}\right)^6\right],
\end{equation}
where $ij$ denotes the particle pair, $\varepsilon$ and $\sigma$ are the model dependent energy and length parameters, respectively.
Clearly this approach is rather elegant.
Knowing the correct values for $\sigma$ and $\varepsilon$, the simulations show good agreement with experiments~\cite{zhangMolecularSimulationThin2019}.
Which is quite impressive if we take into account that the method is considered noisy.

However there is no free lunch. 
Because of the microscopic nature of this method it is compute intensive.
Per definition the method needs to know the phase space coordinates, thus the position and the velocity, of every particle.
While it is possible to arrange this information efficiently, the computation of the pair potential quickly becomes quite resource demanding in terms of computational power.
The results from the computation of pair potentials are forces which act on the particles.
This surmounts to a lot of computational resources for small physical volumes, which is in the context of fluid flows and film formation problematic.
Assuming a box of a few nanometers in all directions and a physical time of about a picosecond ($10^{-12}s$) is already a lot to do for a common laptop.
Clearly a lot is depending on the implementation of the algorithm. 
There are well developed, open source and supported software solutions that not only scale well but have a mature user interface, e.g. LAMMPS and GROMACS~\cite{plimptonFastParallelAlgorithms1995, berendsenGROMACSMessagepassingParallel1995, lindahlGROMACSPackageMolecular2001}.  
However all this development cannot overcome that this method is suited only for relatively small systems and not for macroscopic scales.

Another problem that is inherent to MD simulations is the vast zoo of interaction potentials.
Water for example is not a Lennard-Jones liquid and therefore the potential Eq.~(\ref{eq:Len-Jon}) cannot be used or water-water interactions.
Although water seems to be a simple molecule with just two hydrogen atoms and one oxygen atom it is in fact a rather complex substances to simulate.
There exists not a single model for how water should interact but several, all with very fine tuned adjustments.
Some include for example the polarity of the water molecule while others neglect it.
Without knowing why a certain water model is more suited than another one, this approach can come down to a black box.
This method therefore requires a lot of literature knowledge and validation with both theory and where possible with experiments.
That said apart from Zhang et al.~\cite{zhangMolecularSimulationThin2019} MD has been used on a diverse set of thin film related problems such as pearling~\cite{koplikPearlingInstabilityNanoscale2006}, dewetting~\cite{bertrandDynamicsDewettingNanoscale2007} or dynamics of droplets~\cite{liuActuatingWaterDroplets2015, wangMolecularDynamicsSimulations2015} to name just a few.

>>>>

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes.}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we cannot find an analytical solution to every given flow problem, which is due to the shape and the structure of the Navier-Stokes equations.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille flow'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{suteraHistoryPoiseuilleLaw}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number\footnote{The Stokes number is a dimensionless number that defines the behaviour of particles suspended in a flow and can be calculated with $St = \frac{t_0u_0}{l_0}$, where $t_0$ is a relaxation time of the particles, $u_0$ is a characteristic velocity and $l_0$ is a characteristic length scale.}.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
However, not all flows are laminar and in fact most natural flow admit some turbulent behaviour. 
Understanding flows with high Reynolds numbers still poses a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Several physics problems are being studied to a large extent with numerical tools, e.g., quantum field theory~\cite{montvayQuantumFieldsLattice1994}, solid mechanics~\cite{cardiffThirtyYearsFinite2021}, and as outlined in the introduction Chap.~\ref{chapter:intro}, computational fluid dynamics (CFD) is one of those fields.
Although chip manufacturers have a hard time to keep up with Moore's law, computing power is still growing rapidly~\cite{schallerMooreLawPresent1997}.
Instead of ever faster processors (measured in clock rate), the trend is shifting in the direction of parallel and accelerated computing. 
Throughout the years several accelerator designs emerged.  
One chip design that works particularly well as accelerators are graphics processing units (GPUs).
While a GPU lacks the complexity of a central processing unit (CPU) it excels at doing simple tasks over and over again.
As the name suggests they were initially developed to generate a graphic signal that could be read by a screen. 
It turns out that several algorithms developed for solving differential equations are well suited to be calculated on the GPU. 
Among them is the lattice Boltzmann method, at least a very basic implementation of the method.
GPUs however are not the only devices to be used as accelerators.
Intel for example used a different approach to accelerator computing and already some time ago developed the Xenon Phi.
However with lack of performance as compared to Nvidia's graphics cards and company intern decisions Intel ended the Xenon Phi project in 2020.
The successor of the Xenon Phi project is the very recent \textit{oneAPI} approach and its hardware~\cite{christgauPortingLegacyCUDA2020}.
Shifting the focus towards heterogeneous compute architectures, or as Rick Stevens\footnote{Rick Stevens is Professor at the Department of Computer Science and Associate Laboratory Director, Computing, Environment and Life Sciences at Argonne National Laboratory.} said: ''The future of advanced computing requires heterogeneous hardware to maximize the computing power needed for exascale-class workloads. The oneAPI industry initiative Intel is spearheading will ensure that programming across diverse compute architectures is greatly simplified.''
Not only is this Intel's idea but a general trend in high performance computing (HPC) moving from CPU only clusters to heterogeneous hardware.
Apart from Intel more and more companies joined the development of workload optimized hardware.
Which is possible not only but also due to the technological inventions from ARM (Advanced RISC Machines) and other system on a chip (SoC) design companies.
SoC designs optimized for specific workloads are the key business of ARM, which is a paradigm change of the industry.
For the longest time established chip manufactures also defined the designs of a chips.
Today SoC designs can be manufactured at project manufacturers with TSMC (Taiwan Semiconductor Manufacturing Company Limited) being one of them.
TSMC is also the world leader in the production of semiconductors for computer chips. 
They moved the resolution limit for single transistors from 12nm to about 5nm in roughly 5 years.
These technological steps allowed on the other hand companies such as Apple to design SoCs for their specific needs. 
The M1 Max is from this perspective a very interesting device, although it is not an x86 design.

@@

\section{Numerical methods}
\label{sec:numerical_methods}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{graphics/800px-Finite_Differences.svg.png}
    \caption{Discretization of a continuous function y (in red) along a discrete space x ($x_0, x_1, x_2$,...).
    Consecutive points on the x-axis are separated by a uniform distance $h$.
    }
    \label{fig:finite_difference}
\end{figure}
The lattice Boltzmann method is one among many to simulate the behaviour of a fluid.
Especially when it comes to numerical simulations of the thin film equation, see Eq.~(\ref{eq:simple_thin_film}), other numerical approaches are more established and have been successfully applied for more than thirty years~\cite{beckerComplexDewettingScenarios2003, peschkaSignaturesSlipDewetting2019, davidovitchSpreadingViscousFluid2005, meckeThermalFluctuationsThin2005, diezGlobalModelsMoving2000, schwartzSimulationDropletMotion1998}.
The thin film equation is a non-linear (stiff) fourth order PDE which requires choosing an appropriate scheme, because not every numerical scheme will be able to provide first stability and second a correct approximation of the solution.
A numerical scheme is a blue print of how to discretize derivatives and arrange terms such that an iterative process converges towards a solution.
That said, developing new numerical schemes for PDEs is still a very active research area.
One way to compute this differential operators on a computer is via finite differences.

\subsection{Finite differences}
Using finite differences means nothing more than discretizing the continuous problem into small (infinitesimal) discrete segments.
Figure~\ref{fig:finite_difference} illustrates a discretization approach where the function $f(x)$ is approximated by a finite set of points which are shown as black dots on the red line.
At every point $x_i$ along the x-axis we know the value of the function $f(x_i)$.
This however does not mean that we know the value of $f(x)$ at arbitrary $x$ values and even worse we do not know $\partial_x f(x_i)$ at $x_i$. 
For a continuous function, e.g., the film thickness $h(\mathbf{x},t)$, we can analytically compute the derivatives. 
Here we are left with a discrete set of points and have yet to come up with an idea to calculate $\nabla h(\mathbf{x}_i, t)$, which is crucial to compute the film pressure, Eq.~(\ref{eq:pfilm}).

Discretization of a derivative is a long studied problem~\cite{booleTreatiseCalculusFinite1872, jordanCalculusFiniteDifferences1965}.
The Taylor expansion gives us a tool to approximate the solution to a differential.
In one spatial dimension there are three different methods to compute the derivative, neglecting boundary conditions there is
\begin{equation}\label{eq:forward_dif}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_i)}{h},
\end{equation}
where $h$ is the spacing between two consecutive points as introduced in Fig.~\ref{fig:finite_difference}. 
Subtracting the value of the function $f(x_{i+1})$ from $f(x_i)$ is called forward difference.
Changing the interval the backward difference is given as
\begin{equation}\label{eq:backward_dif}
    \partial_x f(x_i) = \frac{f(x_i) - f(x_{i-1})}{h}.
\end{equation}
Both the forward as well as the backward differential are first order in their error, beyond first order is the central difference approach
\begin{equation}\label{eq:center_diff}
    \partial_x f(x_i) = \frac{f(x_{i+1}) - f(x_{i-1})}{2h},
\end{equation}
where the value of the derivative is averaged over a larger interval. 
The second derivative is then simply applying the differential operator (forward, backward, central) to  the first derivative. 
In case of central differences this yields
\begin{equation}\label{eq:second_central_diff}
    \partial_x^2 f(x_i) = \frac{f(x_{i+1})-2f(x_i)+f(x_{i-1})}{^2}.
\end{equation}

However all finite difference schemes suffer from the concrete choice of the discretization, which is often referred to as discretization artifacts.
The starting equation is usually a continuous one in both space and time, but the numerical solver requires a discretized system. 
This can be done as shown in Fig.~\ref{fig:finite_difference} by introducing a grid for the spatial dimensions. 
For now we assume that we can do the same with time and put the temporal dimension into a discrete grid. 
If we take a sine wave and discretize it with an arbitrary grid spacing $h$, that grid spacing will introduce an error to our finite difference scheme. 
We mark two points along the curve and connect them, which yields a straight line. 
Therefore all information in between the points is lost, instead of $y(x) = \sin(x)$ one would guess $y(x) = kx + d$ for the points.
Besides the fact that information about the function is lost it still could be a good enough approximation for the derivative, especially if the points are not too far apart.
Increasing the number of points along the curve, thus making $h$ smaller, will increase the accuracy of the resulting approximation as displayed in Fig.~\ref{fig:finite_difference}.
A good approximation for $y(x) = \sin(x)$ requires therefore more than just two points.
Which explains the so called fidelity of the finite difference method~\cite{guruswamyReviewNumericalFluids2002}.
One needs a fine enough grid to resolve the to be studied dynamics, but the smaller the spacing $h$ gets the more numerically demanding is the calculation.

In fact there are further limitations to the problem of numerically approximating a PDE.
Numerical methods for PDEs, e.g., the thin film or the wave equation, need to satisfy the CFL condition~\cite{courantUeberPartiellenDifferenzengleichungen1928}.  
Beyond other things it states that choosing an increment for time cannot be independent of the spatial resolution.
This condition can be formalized as
\begin{equation}\label{eq:CFL}
    C = \Delta t \left(\sum_{i=1}^n \frac{u_{x_i}}{\Delta x_i}\right) \leq C_{max},
\end{equation}
with $\Delta x_i$ and $\Delta t$ being the spatial and temporal increment, respectively and $u_{x_i}$ being the magnitude of the velocity in the respective direction.
In case of Fig.~\ref{fig:finite_difference} we have $\Delta x = h$. 
Behind this concept is the idea that propagation of information has an upper limit.
For fluid dynamics problems this upper limit is the speed of sound, in electrodynamics it would be the speed of light.
Any propagation larger than this upper limit introduces stability issues.
These issues are purely numerical as we know that air can flow with velocities larger than the speed of sound, however by doing so the flow creates a so called shock.

Thinking of a wave package, numerically it should be ensured that several time iterations are computed before the package has reached the next spatial point of the grid.
The maximal value that the Courant number can reach for an explicit solver, to which for example the lattice Boltzmann method belongs (assuming a BGK collision operator), is $C_{max} = 1$.
For implicit methods the maximal Courant number can be larger than one.
The iteration from $u(x_i,t_0)$ to $u(x_i,t_1)$ is called a time step. 
The ``distance'' between consecutive time values, here $\Delta t = t_1 - t_0$ must ensure that forces and therefore acceleration are well within the stability condition Eq.~(\ref{eq:CFL}).

@@

\subsection{Spectral methods}
Another strategy to solve a differential equation is to perform a transformation.
One rather famous transformation is the Fourier Transformation (FT), which we have used to find the dispersion relation in Sec.~\ref{sec:thin_films} and will encounter again in Chap.~\ref{chapter:second_paper}.
The set of solvers using a FT to solve a given differential equation is called spectral methods.
Doing so has the advantage of interchanging the derivatives with simple multiplications.
For example the term $\partial_x f(x)$ transforms to  
\begin{equation}\label{eq:fourier_transform}
    \partial_x f(x) = q\cdot \tilde{f}(q),
\end{equation}
where $\tilde{f}(q)$ is the Fourier transformed of $f(x)$, e.g., $\tilde{f}(q) = \delta(q - a/2\pi)$ for $f(x) = e^{iax}$. 
Instead of solving a differential equation one solves an algebraic equation, or a set of algebraic equations.
While it is often straightforward to perform the transformation it can become problematic to perform the backward transformation. 
Usually the expressions are fairly complicated and often the backward transformation into real space is not easy to perform.
In Chap.~\ref{chapter:second_paper} we arrive at such an expression for the structure factor $\mathcal{S}(q, t)$.
A structure factor is a measure of the distribution of length scales, roughly speaking.
In this specific case there is no need for a backward transformation because (scattering) experiments usually supply data in Fourier space.

However, when approximating the thin film equation, by iteratively solving an equation, we may be interested in forces, pressures or simply the thickness $h(\mathbf{x},t)$, which is why we need a backward transformation.
Numerically speaking the operators for transformations are matrices, which depend on the specific problem.
Solutions in real space are generated by inversion of these operators and therefore by inversion of an (arbitrary) matrix.
Computing the inverse of a matrix is a demanding problem. 
Nevertheless, these transformation approaches are often quite helpful finding an exact solution.
One rather prominent example is the computation of Green's functions, e.g., in electrostatics~\cite{greenEssayApplicationMathematical1889}.
For the shallow water theory, see Sec.~\ref{sec:theory_shallow_water}, one class of effective methods are the (discontinuous) Galerkin methods~\cite{ernTheoryPracticeFinite2004} and to some extent the related finite element method (FEM).

Literature concerning simulations of the thin film equation using spectral solvers is rather sparse. 
That does however not mean that spectral methods are ill suited for this specific problem.
For example, Dur{\'a}n-Olivencia, M. et al. used a spectral method to approximate the thin film equation. 
Their work also introduced a self-consistent approach to tread thermal fluctuations~\cite{duran-olivenciaInstabilityRuptureFluctuations2019}.

\subsection{Molecular dynamics}
The third approach we discuss in this section is the molecular dynamics method (MD)~\cite{haileMolecularDynamicsSimulation1997, zhangMolecularSimulationThin2019, wengMolecularDynamicsInvestigation2000}.
When we solve the equations of motion of molecules we do not per se think of a \textbf{thin} film flow, however molecular dynamics can be very insightful when dealing with thin film problems. 
In Sec.~\ref{sec:thin_films} we introduced the term wetting and the three phase contact line.
Advancement of the three phase contact line can be boiled down to the interaction between fluid molecules and substrate molecules, therefore MD is perfectly suited to study e.g., slip~\cite{jabbarzadehWallSlipMolecular1999, segaRegularizationSlipLength2013}. 
Instead of working with macroscopic fields, such as density $\rho$ or velocity $\mathbf{u}$, the method uses a bottom up approach.
The main idea is to compute the trajectory of every particle using Newton's equation of motion, therefore to assume that particles are classical objects.
Interactions as well as boundary conditions can be added via pair potentials.
A common interaction potential is the Lennard-Jones potential
\begin{equation}\label{eq:Len-Jon}
    U(r_{ij}) = 4\varepsilon\left[\left(\frac{\sigma}{r_{ij}}\right)^{12}-\left(\frac{\sigma}{r_{ij}}\right)^6\right],
\end{equation}
where $ij$ denotes the particle pair, $\varepsilon$ and $\sigma$ are the model dependent energy and length parameters, respectively.
Clearly this approach is rather elegant.
Knowing the correct values for $\sigma$ and $\varepsilon$, the simulations show good agreement with experiments~\cite{zhangMolecularSimulationThin2019}.

However, there is no free lunch. 
The method is inherently microscopic which makes it compute intensive.
To compute a particle trajectory we need to know where the particle is and how fast it is, thus its position and the velocity.
In a MD simulation we need to know this information for every particle.
While it is possible to arrange this information efficiently, the computation of the pair potential quickly becomes quite resource demanding in terms of computational power.
The results from the computation of pair potentials are forces which act on the particles.
This surmounts to a lot of computational resources for small physical volumes, which is in the context of fluid flows and film formation problematic.
Assuming a box of a few nanometers in all directions and a physical time of about a picosecond ($10^{-12}s$) is already a lot to do for a common laptop.
Clearly a lot is depending on the implementation of the algorithm. 
There are well developed, open source and supported software solutions that not only scale well but have a mature user interface, e.g., LAMMPS and GROMACS~\cite{plimptonFastParallelAlgorithms1995, berendsenGROMACSMessagepassingParallel1995, lindahlGROMACSPackageMolecular2001}.  
However all this development cannot overcome that this method is suited only for relatively small systems and not for macroscopic scales.

Another problem that is inherent to MD simulations is the vast zoo of interaction potentials.
Water for example is not a simple Lennard-Jones liquid, such as liquid Argon. 
Therefore the potential Eq.~(\ref{eq:Len-Jon}) should not be used for water-water interactions.
Although water seems to be a simple molecule with just two hydrogen atoms and one oxygen atom it is in fact a rather complex substance to simulate.
There exists not a single model for how water should interact but several, all with very fine tuned adjustments.
Some include for example the polarity of the water molecule while others neglect it.
Without knowing why a certain water model is more suited than another one, this approach can come down to a black box.
This method therefore requires a lot of literature knowledge and validation with both theory and where possible with experiments.
That said apart from Zhang et al.~\cite{zhangMolecularSimulationThin2019} MD has been used on a diverse set of thin film related problems such as pearling~\cite{koplikPearlingInstabilityNanoscale2006}, dewetting~\cite{bertrandDynamicsDewettingNanoscale2007} or dynamics of droplets~\cite{liuActuatingWaterDroplets2015, wangMolecularDynamicsSimulations2015} to name just a few.
